---
title: "top_PICS_distribution"
output: html_document
date: "2024-10-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(readxl)
library(ggplot2)
mpra_pics <-read_excel(paste0("~/Desktop/Ho et al. writings/Code_from_scratch/data/240914_Ho_et_al_tables (Max's Version).xlsx"),sheet=6)
```

Reviewer #3 "For the enrichment analyses in lines 102-114, it would be helpful for the authors to share the distribution of highest posterior probability for each variant in the MPRA library. Based on the way the library was described (i.e., identifying highly associated autoimmune risk variants and choosing variants in LD with those tag variants), it is surprising that there would be sufficient variants with low posterior probability in the library to make a conclusion that variants that were emVars had higher posterior probabilities."

```{r}
ggplot(data=mpra_pics, aes(x=top_pics)) + geom_histogram(fill="darkorange") + theme_bw() + labs(title="Distributon of top PICS posterior probablilties")
max(mpra_pics$top_pics, na.rm = T)
min(mpra_pics$top_pics, na.rm = T)
mean(mpra_pics$top_pics, na.rm = T)
median(mpra_pics$top_pics, na.rm = T)
```

Here it is clear that the distribution of pics values is skewd towards zero and there are only a few variants with pics > 0.5

