---
title: "Final motifbreakr enrichment analysis"
author: Max Dippel
date: "2024-04-29"
output:
  html_document:
    toc: true
    toc_float: true
---

This is the motifbreakr analysis for "Genetic and epigenetic screens in human primary T cells link candidate causal autoimmune variants to T cell proliferation" (Ho et al. *in review*). 

Steps to motifbreakr analysis:

1. Create a Granges bed file of the variants mpra tested in the MPRA

2. Run motifbreakR function to generate TF binding data on the MPRA variants

3. Merge motifbreakr and MPRA data

4. Run t-test of primary T cell MPRA expression of variants which do and do not bind to each tf 

5. Repeat step 4 with jurkat mpra expression data

6. Run t-test analysis for variants fine mapped to each disease

7. Merge the primary tcell and unstimulated jurkat data

8. Compare the results of jurkat and primary T cells

Loading packages
```{r packages,message=FALSE,warning=FALSE}
library(stringr)
library(data.table)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(SNPlocs.Hsapiens.dbSNP155.GRCh38)
library(tidyverse) 
library(readxl)
library(rsample) 
# library(ggpubr) 
library(cluster)    
#library(factoextra) 
library(dendextend)
library(rjson)
#library(MotifDb)
library(motifbreakR)
#library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg38)
#library(BSgenome.Hsapiens.UCSC.hg19)
# library(GenomicRanges)
library(ggdendro)
library(gridExtra)
#library(GenomicFeatures)
```

List directories
```{r}
data.dir <- "~/Desktop/Ho et al. writings/Code_from_scratch2/data/"
plots.dir <- "/Users/mdippel/Desktop/Ho\ et\ al.\ writings/Code_from_scratch2/plots/"
mpra.name1 <- "20241111_tcell"
mpra.name2 <- "20241111_jurkat"
motifbreakr.name <- "2023.11.15.hg38.tcells.glm.all.snps.hocomoco.bed"
```


## Motifbreakr bed file

Create bed file from the MPRA data and create a SNP granges object from them
```{r , eval=FALSE}
# Load MPRA data
tcell_mpra <- read.table(paste0(data.dir,mpra.name1,"_mpra_merge_hg38_all.txt"), sep="\t", header=T)

# subset the mpra to only the columns you seed
mpra.bed<-subset(tcell_mpra, select=c(chr, snp_start, snp_end, SNP))
# Create a name column in the mpra which uses the chromosome, end snp, and SNP to create a name
mpra.bed$name<-paste(mpra.bed$chr, mpra.bed$snp_end, str_split_fixed(mpra.bed$SNP, "\\:", 4)[,3], str_split_fixed(mpra.bed$SNP, "\\:", 4)[,4],sep=":")
# create a blank mpra.bed score column
mpra.bed$score<-0
# Create a strand column with just +
mpra.bed$strand<-"+"
# Now subset even further to only the colums you need
mpra.bed<-subset(mpra.bed, select=c(chr, snp_start, snp_end, name,score, strand))
# create the mpra bed file
write.table(mpra.bed, paste0(data.dir,"motifbreakr/tcell_mpra_snps.bed"), row.names = F, col.names = F, sep="\t", quote=F)
# Create SNPs GRanges object
tcell.snps.hg38.frombed <- snps.from.file(file = paste0(data.dir,"motifbreakr/tcell_mpra_snps.bed"),
                                  dbSNP = SNPlocs.Hsapiens.dbSNP155.GRCh38,
                                  search.genome = BSgenome.Hsapiens.UCSC.hg38,
                                  format = "bed")
```


## Motifbreakr function

Run motifbreakr with Hocomoco v.11 and the bed file
```{r motifbreakr function, eval=FALSE}
tcells.hocomoco.bed.results <- motifbreakR(snpList = tcell.snps.hg38.frombed, filterp = TRUE,
                       pwmList = subset(MotifDb, 
                                        dataSource %in% c("HOCOMOCOv11-core-A", "HOCOMOCOv11-core-B", "HOCOMOCOv11-core-C")),
                       threshold = 1e-4,
                       method = "log",
                       bkg = c(A=0.25, C=0.25, G=0.25, T=0.25),
                       BPPARAM = BiocParallel::SerialParam())

results <- data.frame(tcells.hocomoco.bed.results, stringsAsFactors = F)
results$seqnames <- as.character(results$seqnames)
results$motifPos <- as.character(results$motifPos)
write.table(results,paste0(data.dir,"motifbreakr/",motifbreakr.name,".motifbreakr.results.txt"), row.names=FALSE,col.names=TRUE,sep="\t")
```

## Primary tcell analysis

### Set up 

Set up Primary t cell  MPRA and the motifbreakr results 
```{r primary tcell set up}
# Set up cell type
celltype <- "Primary Tcell"
celltype_dir <- "tcell"
rna_cell.type <- "tcell"
# Set up motif db option
motif_db <- "hocomoco"
# Set up directories for tables
mpra.dir.1 <- paste0(data.dir, mpra.name1,"_mpra_merge_hg38_filtered.txt")
motifbreakr_dir <- paste0(data.dir,"motifbreakr/",motifbreakr.name,".motifbreakr.results.txt")
# Give the cell type a distinct color
celltype_color <- "#E14A4A"
```

# import RNA-seq data
```{r}
################
# Import RNA-seq data
################
rna_cell.type 
# Importing dice gene expression data for primary CD4 tcells
# https://dice-database.org/downloads
tcell_TPM <- read.csv(paste0(data.dir,"motifbreakr/","CD4_NAIVE_TPM.csv"), header=T)
# Only the max TPM values
tcell_TPM <- subset(tcell_TPM, select=c("Additional_annotations", "X0"))
# Write the correct gene name
tcell_TPM$Additional_annotations <- gsub(";protein_coding","",tcell_TPM$Additional_annotations)
# Filter for over 1 TPM
tcell_TPM <- subset(tcell_TPM, X0 >= 1)
# Rename the columns to merge with the motif data
names(tcell_TPM) <- c("tf","TPM")

jurkatTPM <- read.table(paste0(data.dir,"motifbreakr/","GSE145453_norm_counts_TPM_GRCh38.p13_NCBI.tsv"), sep = '\t', header=T)
#https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE145453
# Resting Jurkat GSM4318505, GSM4318506
# Stimulated Jurkat GSM4318507, GSM4318508
# Stimulated + chemokine Jurkat GSM4318509, GSM4318510
# The annotate file in the GEO data set Human.GRCh38.p13.annot.tsv did not have that many genes (39000 our of 20000) so I am going to use an online tool to convert the gene IDs to 
# https://www.syngoportal.org/convert
# This takes a long time but the results will be worth it. Use NCBI entrez Gene (need to confirm)
annotateTPM <- read_excel(paste0(data.dir,"motifbreakr/","idmap.xlsx"))
annotateTPM <- subset(annotateTPM, select=c(query,symbol))
names(annotateTPM) <- c("GeneID","tf")
jurkatTPM <- merge(jurkatTPM,annotateTPM, by="GeneID")
jurkatTPM$unstim_jurkat_TPM <- pmax(jurkatTPM$GSM4318505,jurkatTPM$GSM4318506)
jurkatTPM$stim_jurkat_TPM <- pmax(jurkatTPM$GSM4318507,jurkatTPM$GSM4318508)
unstim_jurkat_TPM<- subset(jurkatTPM, select=c(tf,unstim_jurkat_TPM))
unstim_jurkat_TPM <- subset(unstim_jurkat_TPM, unstim_jurkat_TPM >= 1)
stim_jurkat_TPM<- subset(jurkatTPM, select=c(tf,stim_jurkat_TPM))
stim_jurkat_TPM <- subset(stim_jurkat_TPM, stim_jurkat_TPM >= 1)
```

### import and reformat data
Import motifbreakr and mpra tables and modify them to be used in this analysis
```{r primary t cell import data}
#########################
# Creating a data set with motif data and MPRA data
#########################

# Import motif data
motif.dat <- read.table(motifbreakr_dir, sep="\t", header=T)

# Import  MPRA
mpra <- read.table(mpra.dir.1, sep="\t", header=T)

# Makes geneSymbol (TF name) all uppercase
motif.dat$geneSymbol <- toupper(motif.dat$geneSymbol)
# make geneSymbol (TF Name) change from slash / to underscore so that the plots can save
motif.dat$geneSymbol <- gsub("/","_",motif.dat$geneSymbol)

# Filter by genes seen in RNAseq data (hash out the line below to eliminate the RNAseq filter). This is now going to be applied downstream after the t-test data
# motif.dat <- merge(motif.dat,get(paste0(rna_cell.type,"_TPM")),by="geneSymbol")

# Subset the motif data
motif.dat<-subset(motif.dat,  select=c(seqnames, end, REF, ALT,SNP_id, geneSymbol,scoreRef ,scoreAlt, alleleDiff))
# Subset MPRA
mpra <- subset(mpra, select=c(SNP,A.log2FC,B.log2FC,LogSkew,mpra_sig,rsid,ref_allele,alt_allele))

######## code for merging mpra and motifbreakr by chromosome and position ########
# Get rid of the chr in motifbreakr SNP_id column to conform with mpra SNP column
motif.dat$SNP <- gsub("chr","",motif.dat$SNP_id)
# Merge Motifbreakr and MPRA ( we expect a similar number to the 61570 that are in the motif.dat right now (example run not represenative of all runs))
motif.mpra.dat<-merge(motif.dat, mpra, by="SNP", all.x=T, all.y=T)
# 61579. Probably due to mpra duplicates. 
# An issue here is that there are variants in MPRA with SNP names with multiple alternate alleles, but that doesn't seem to be a huge issue. 

####### code for merging by rsid which is not used anymore #########
# Create a SNP column for the motifbreakr data (used to merge with MPRA)
# motif.dat$rsid<- motif.dat$SNP_id
# Merge Motifbreakr and MPRA
# motif.mpra.dat<-merge(motif.dat, mpra, by="rsid", all.x=T, all.y=F)

# Make sure mpra data is only the mpra_sig categories that we want
motif.mpra.dat <- subset(motif.mpra.dat, mpra_sig %in% c("Enhancer_Skew","Enhancer_nSkew","nEnhancer_nSkew"))
# Create emVar not emVar column
motif.mpra.dat$variant_skew <- ifelse(motif.mpra.dat$mpra_sig == "Enhancer_Skew", "emVar", "non_emVar")

# Make sure alleles for Motifbreakr and MPRA are the same 
# Rename the columns 
motif.mpra.dat$motifbreakr_ref <- motif.mpra.dat$REF
motif.mpra.dat$mpra_ref <- motif.mpra.dat$ref_allele
# Create a column with a 1 when the motifbreakr and mpra alleles match (Improve this loop)
motif.mpra.dat$ref_agree <- NA
motif.mpra.dat$ref_agree <- as.integer(motif.mpra.dat$motifbreakr_ref == motif.mpra.dat$mpra_ref)
# Subset to only variants in which the reference allele in the MPRA and motifbreakr match
motif.mpra.dat <- subset(motif.mpra.dat, ref_agree==1)

# Make sure alleles for Motifbreakr and MPRA are the same 
# Rename the columns 
motif.mpra.dat$motifbreakr_alt <- motif.mpra.dat$ALT
motif.mpra.dat$mpra_alt <- motif.mpra.dat$alt_allele
# Create a column with a 1 when the motifbreakr and mpra alleles match (Improve this loop)
motif.mpra.dat$alt_agree <- NA
motif.mpra.dat$alt_agree <- as.integer(motif.mpra.dat$motifbreakr_alt == motif.mpra.dat$mpra_alt)
# Subset to only variants in which the alternate allele in the MPRA and motifbreakr match
motif.mpra.dat <- subset(motif.mpra.dat, alt_agree==1)

# Add back in the variants without motifbreakr alleles
motif.mpra.dat<-merge(motif.mpra.dat, mpra, by=names(mpra), all.x=T, all.y=T)

# Generate unique MPRA sites 
motif.mpra.dat$unique_snp_tf <- paste0(motif.mpra.dat$seqnames,":",motif.mpra.dat$end,":",
                                    motif.mpra.dat$REF,":",motif.mpra.dat$ALT,"_",motif.mpra.dat$geneSymbol)
motif.mpra.dat<-motif.mpra.dat[order(-abs(motif.mpra.dat$alleleDiff)),]
motif.mpra.dat<-motif.mpra.dat[!duplicated(motif.mpra.dat$unique_snp_tf),]

write.table(motif.mpra.dat, paste0(data.dir,"motifbreakr/tcell/", "motif.mpra.dat_",celltype_dir,"_",motif_db,".txt"), row.names=FALSE,col.names=TRUE,sep="\t", quote=F)


TF_list <-  unique(motif.mpra.dat$geneSymbol)

##################

# Plots which flip the sign of logskew for negative binders and using a binary readout for all differential TF binding

##################

# Create the TF bind or not bind data with a column for LogSkew, SNP and a column for each TF with the value of alleleDiff inside it if the TF binds to that particular row's SNP

TF_all_bind_data <- subset(motif.mpra.dat, select=c(geneSymbol,LogSkew,SNP,alleleDiff))
# Pivot the data to wide format using tidyr::pivot_wider
TF_all_bind_data <- TF_all_bind_data %>%
  pivot_wider(names_from = geneSymbol, values_from = alleleDiff)

TF_all_bind_data <- unique(TF_all_bind_data)

# The filter for the number of observations needed to have sufficient sample size for each TF
TF_all_bind_data <- TF_all_bind_data[, (colSums(!is.na(TF_all_bind_data)) >= 5)==TRUE]
# If you combine this with the activity filter the results really change a lot

# For each MPRA variant, find the fold change of the allele with the largest absolute fold change
# This creates a column full of NAs for the loop to fill
mpra$max_abs_fc <- NA
# This loop calculates the column of max_abs_fc. It takes the two alleles A and B and calculates which one's absolute value is larger. That larger value is the value of the new column (can be positive or negative). 
for(i in 1:nrow(mpra)){
  if(abs(mpra[i,]$A.log2FC)>abs(mpra[i,]$B.log2FC)){
    mpra[i,]$max_abs_fc<-mpra[i,]$A.log2FC
  }else{
    mpra[i,]$max_abs_fc<-mpra[i,]$B.log2FC
  }
}

# Add back in the varaints which bind to no TFs
logskew_fc_all_vars <- subset(mpra, select=c(SNP,LogSkew,max_abs_fc))

TF_all_bind_data <- merge(logskew_fc_all_vars,TF_all_bind_data, by=c("SNP","LogSkew"), all.x=T)

write.table(TF_all_bind_data, paste0(data.dir, "motifbreakr/tcell/", "TF_all_bind_data_",celltype_dir,"_",motif_db,".txt"), row.names=FALSE,col.names=TRUE,sep="\t", quote=F)
```

### analysis loop
Motifbreakr analysis loop for primary t cells
```{r primary t cell analysis loop}
# TF list 
TF_list <- names(TF_all_bind_data)
TF_list <- TF_list[4:length(TF_list)] 

# Create test with different samples of flipping the same amount of variants in the no bind as you do in the yes bind.
# Now create a data frame with p value and effect size for all the TFs. 
t.test.all.bind.motif.dat <- NULL
hundred.p.values.per.tf <- data.frame(p.val=1:100)
# T-test to test whether the distribution of LogSkew differs between variants that bind to a TF and those that don't bind 
for(tf in TF_list){

temp_data<-TF_all_bind_data
# How many variants which bind to this particular tf are being flipped?
# Bind to TF 
specific_tf_data <- subset(temp_data,  select=c("SNP","LogSkew","max_abs_fc",tf))
# This is the sample size of the allele which are flipped for the tfs which bind to a particular TF. Change this sign as well as the one below to flip the plot
size_flipped_allelediff <- nrow(subset(specific_tf_data,  is.na(specific_tf_data[[tf]])==FALSE & specific_tf_data[[tf]]>=0))

# Find the variants which bind to a particular tf
yes_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])!=TRUE)
# flip the signs of logskew for tfs that bind and tfs that don't no bind (which we don't even normally flip) separately. Change the sign with the one above in order to flip the plot
yes_tf_bind$LogSkew <- ifelse(!is.na(yes_tf_bind[[tf]]) & yes_tf_bind[[tf]] >= 0, -yes_tf_bind$LogSkew, yes_tf_bind$LogSkew)
# Find the tfs that do not bind to a particular tf
no_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])==TRUE)
# Find the size of rows that must be flipped by calculating the percent of rows that were flipped from the variants which do bind to a particular tf and multiplying it by the rows of the variants that do not bind 
size_flip_no_bind <- nrow(no_tf_bind)* (size_flipped_allelediff/nrow(yes_tf_bind))
# Pick a certain amount of rows to flip from those TFs which do not bind. 
rows_to_flip <- sample(1:nrow(no_tf_bind), size=size_flip_no_bind, replace=FALSE)
# Now flip the LogSkew of the rows which I picked for the variants which do not bind to that TF
no_tf_bind$LogSkew[rows_to_flip] <- -no_tf_bind$LogSkew[rows_to_flip]

# Number of variants which bind to this particular tf
n <- length(yes_tf_bind$LogSkew)
# Calculating effect size (Cohen's D)
d <- (mean(yes_tf_bind$LogSkew, na.rm = TRUE) - mean(no_tf_bind$LogSkew, na.rm = TRUE)) / (sqrt(((sd(yes_tf_bind$LogSkew, na.rm = TRUE)^2)+(sd(no_tf_bind$LogSkew, na.rm = TRUE)^2)/2)))
mu1 <- mean(yes_tf_bind$LogSkew, na.rm = TRUE)
sd1 <- sd(yes_tf_bind$LogSkew, na.rm = TRUE)
mu0 <- mean(no_tf_bind$LogSkew, na.rm = TRUE)
sd0 <- sd(no_tf_bind$LogSkew, na.rm = TRUE)

# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
skew.bind.t.test <- t.test(x=yes_tf_bind$LogSkew, y= no_tf_bind$LogSkew,alternative = c("two.sided"))
t <- skew.bind.t.test$statistic
df <- skew.bind.t.test$parameter
p <- skew.bind.t.test$p.value

t.test.all.bind.motif.dat <- rbind(t.test.all.bind.motif.dat, data.frame(tf,p,d,n,t,df,mu1,sd1,mu0,sd0))
t.test.all.bind.motif.dat

# Repeat t-test 100 times
#################################
for(i in 1:100){

# How many variants which bind to this particular tf are being flipped?
# Bind to TF 
specific_tf_data <- subset(temp_data,  select=c("SNP","LogSkew","max_abs_fc",tf))
# This is the sample size of the allele which are flipped for the tfs which bind to a particular TF. Change this sign as well as the one below to flip the plot
size_flipped_allelediff <- nrow(subset(specific_tf_data,  is.na(specific_tf_data[[tf]])==FALSE & specific_tf_data[[tf]]>=0))

# Find the variants which bind to a particular tf
yes_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])!=TRUE)
# flip the signs of logskew for tfs that bind and tfs that don't no bind (which we don't even nromally flip) separately. Change the sign with the one above in order to flip the plot
yes_tf_bind$LogSkew <- ifelse(!is.na(yes_tf_bind[[tf]]) & yes_tf_bind[[tf]] >= 0, -yes_tf_bind$LogSkew, yes_tf_bind$LogSkew)
# Find the tfs that do not bind to a particular tf
no_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])==TRUE)
# Find the size of rows that must be flipped by calculating the percent of rows that were flipped from the variants which do bind to a particular tf and multiplying it by the rows of the variants that do not bind 
size_flip_no_bind <- nrow(no_tf_bind)* (size_flipped_allelediff/nrow(yes_tf_bind))
# Pick a certain amount of rows to flip from those TFs which do not bind. 
rows_to_flip <- sample(1:nrow(no_tf_bind), size=size_flip_no_bind, replace=FALSE)
# Now flip the LogSkew of the rows which I picked for the variants which do not bind to that TF
no_tf_bind$LogSkew[rows_to_flip] <- -no_tf_bind$LogSkew[rows_to_flip]

# Calculating effect size (Cohen's D)
d <- (mean(yes_tf_bind$LogSkew, na.rm = TRUE) - mean(no_tf_bind$LogSkew, na.rm = TRUE)) / (sqrt(((sd(yes_tf_bind$LogSkew, na.rm = TRUE)^2)+(sd(no_tf_bind$LogSkew, na.rm = TRUE)^2)/2)))

# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
skew.bind.t.test <- t.test(x=yes_tf_bind$LogSkew, y= no_tf_bind$LogSkew,alternative = c("two.sided"))
p <- skew.bind.t.test$p.value
hundred.p.values.per.tf[[tf]][i] <- p
}
#################################

 }

# Get a subsetted list of tfs which only have significant p-values
sig_tfs <- data.frame(names(hundred.p.values.per.tf)[2:ncol(hundred.p.values.per.tf)])
names(sig_tfs) <- "tf"
for (i in 1:length(TF_list)) {
sig_tfs$max.p[i] <- max(hundred.p.values.per.tf[[TF_list[[i]]]])
}


t.test.all.bind.motif.dat <- merge(t.test.all.bind.motif.dat,sig_tfs,by="tf")

write.table(t.test.all.bind.motif.dat, paste0(data.dir,"motifbreakr/tcell/", "t.test.all.bind.motif.dat_",celltype_dir,"_",motif_db,".txt"), row.names=FALSE,col.names=TRUE,sep="\t")

t.test.all.bind.motif.dat <- read.table(paste0(data.dir,"motifbreakr/tcell/", "t.test.all.bind.motif.dat_",celltype_dir,"_",motif_db,".txt"),header=TRUE,sep="\t")

# Filter by genes seen in RNAseq data (hash out the line below to eliminate the RNAseq filter)
t.test.all.bind.motif.dat <- merge(t.test.all.bind.motif.dat,get(paste0(rna_cell.type,"_TPM")),by="tf")

top_sample_size <- t.test.all.bind.motif.dat[order(-t.test.all.bind.motif.dat$n),]

GFI1B_sample_size <- subset(t.test.all.bind.motif.dat, tf=="NFKB1")$n

yes_bind_sample_size_hist <- ggplot(data=t.test.all.bind.motif.dat, aes(n)) + 
  geom_histogram(binwidth = 10, fill="#5D93CD") + theme_bw() +labs(title="Sample size of variants which bind to each tf",caption= paste0("top tfs: ", top_sample_size$tf[1], ", ",top_sample_size$tf[2],", ", top_sample_size$tf[3],", ", top_sample_size$tf[4],", ",top_sample_size$tf[5])) + geom_vline(xintercept = GFI1B_sample_size, color="#E14A4A") +annotate("text", x=(GFI1B_sample_size-7), y=16, label="NFKB1", angle=90) + annotate("text", x=(GFI1B_sample_size + 7), y=16, label=GFI1B_sample_size, angle=90)
yes_bind_sample_size_hist

# Lowest number of variants which bind to a particular TF
min(t.test.all.bind.motif.dat$n)
```

### plots 
Motifbreakr analysis plots for primary t cells
```{r primary t cell plots}
# Creating labels for the plot below
t.test.all.bind.motif.dat_ranked <- t.test.all.bind.motif.dat[order(t.test.all.bind.motif.dat$p),]
t.test.all.bind.motif.dat_ranked$rank <- c(1:nrow(t.test.all.bind.motif.dat_ranked))
rows <- 0.15*nrow(t.test.all.bind.motif.dat_ranked)
labels <- t.test.all.bind.motif.dat_ranked$tf[1:rows]

# Plot of t-test p-values and effect size of LogSkew comparing SNPs which bind and do not bind to TF
plot1 <- ggplot(data=t.test.all.bind.motif.dat_ranked , aes(x = d, y=(-log10(p)))) + geom_point(colour = celltype_color) + geom_text_repel(data=subset(t.test.all.bind.motif.dat_ranked , tf==labels), aes(label=tf), max.overlaps = 50) + theme_bw() + xlab("Cohen's D (effect size)") + ylab("T-test log10 p-value") + 
  labs(title="", subtitle = "LogSkew comparing SNPs which bind and do not bind to TF", caption = paste0("Cell type: ", celltype, " ; motif db: ", motif_db)) + 
  theme(panel.grid.minor = element_blank()) + geom_point(data=subset(t.test.all.bind.motif.dat_ranked, max.p>=0.05), colour = "gray50")
print(plot1)

plot.name <- paste0("Volcano_plot_", celltype,"_", motif_db, "_all_flipped_binding", "", ".pdf")
plot.loc <- paste0(plots.dir,"motifbreakr/tcell")
ggsave(plot.name,plot = plot1, path = plot.loc,height=4.5,width=7)


t.test.all.bind.motif.dat_ranked <- t.test.all.bind.motif.dat[order(t.test.all.bind.motif.dat$p),]
top_tf <- t.test.all.bind.motif.dat_ranked$tf[1]
top_ten_tfs <- t.test.all.bind.motif.dat_ranked$tf[1:10]

all.top.tf.dat <- data.frame(SNP=NULL, rsid=NULL,geneSymbol=NULL, alleleDiff=NULL, LogSkew=NULL, mpra_sig=NULL, top_pics=NULL, top_disease=NULL, dhs_Tcell_merged=NULL)

for (tf in top_ten_tfs){
  top.tf.dat <- subset(motif.mpra.dat, geneSymbol==tf, select=c(geneSymbol,alleleDiff,SNP,LogSkew,mpra_sig))

skew_corr <- ggplot(data=top.tf.dat, aes(x=alleleDiff,y=LogSkew)) + geom_point(color=celltype_color) + geom_smooth(method="lm",colour=celltype_color) + theme_bw() +
  labs(title = paste0("Correlation between motifbreakr allelediff and LogSkew for ", tf), 
       subtitle= paste0("Cell type: ", celltype, " ; motif db: ", motif_db), 
       caption = "TF Binding: All TF binding original LogSkew")
print(skew_corr)

plot.name <- paste0("Corr_allelediff_LogSkew_", tf,"_", celltype,"_", motif_db, ".pdf")
plot.loc <- paste0(plots.dir,"motifbreakr/tcell")
ggsave(plot.name,plot = skew_corr, path = plot.loc, width = 7, height = 7)
}
```

## Jurkat analysis

### Set up

Set up Jurkat MPRA and the motifbreakr results 
```{r jurkat set up}
# Do this for another cell type
# Set up cell type
celltype <- "Unstimulated Jurkat"
celltype_dir <- "unstim_jurkat"
rna_cell.type <- "unstim_jurkat"
# Set up directories for tables
mpra.dir.2 <- paste0(data.dir, mpra.name2,"_mpra_merge_hg38_filtered.txt")
motifbreakr_dir <- paste0(data.dir,"motifbreakr/",motifbreakr.name,".motifbreakr.results.txt")
# Set up motif db option
motif_db <- "hocomoco"
celltype_color <- "#5D93CD"
```

### import and reformat data

Import motifbreakR and MPRA tables and reformat them to be used in this analysis
```{r jurkat import data}
#########################
# Creating a data set with motif data and MPRA data
#########################

# Import motif data
motif.dat <- read.table(motifbreakr_dir, sep="\t", header=T)

# Import  MPRA
mpra <- read.table(paste0(mpra.dir.2), sep="\t", header=T)

# eliminate NAs from FC column
#mpra <- subset(mpra, is.na(A.log2FC)==FALSE & is.na(B.log2FC)==FALSE)
# Lose 6 rows in data set
# Make fold change filter with top FC value >1. THIS FILTER WITH SIGNIFICANTLY CHANGE RESULTS. 
#for(c in 1:nrow(mpra)){
#  if(abs(mpra$A.log2FC[c]) >= abs(mpra$B.log2FC[c])){
#    mpra$max.abs.log2FC[c] <- abs(mpra$A.log2FC[c])
#  }else{mpra$max.abs.log2FC[c] <- abs(mpra$B.log2FC[c])
#  }
# }
# Hash out this line of code to remove fold change filter. If you turn this filter on you must also turn the sample size filter on so that there are a sufficent # of observations for each t-test
# mpra <- subset(mpra, max.abs.log2FC >= 1)
# With FC cutoff of 1 only 2192 out of ~18000 variants left!
# With FC cutoff of 0.5 only 6430 out of ~18000 variants left!

# Makes geneSymbol (TF name) all uppercase
motif.dat$geneSymbol <- toupper(motif.dat$geneSymbol)
# make geneSymbol (TF Name) change from slash / to underscore so that the plots can save
motif.dat$geneSymbol <- gsub("/","_",motif.dat$geneSymbol)

# Filter by genes seen in RNAseq data (hash out the line below to eliminate the RNAseq filter). This is now going to be applied downstream after the t-test data
# motif.dat <- merge(motif.dat,get(paste0(rna_cell.type,"_TPM")),by="geneSymbol")

# Subset the motif data
motif.dat<-subset(motif.dat,  select=c(seqnames, end, REF, ALT,SNP_id, geneSymbol,scoreRef ,scoreAlt, alleleDiff))
# Subset MPRA
mpra <- subset(mpra, select=c(SNP,A.log2FC,B.log2FC,LogSkew,mpra_sig,rsid,ref_allele,alt_allele))

######## code for merging mpra and motifbreakr by chromosome and position ########
# Get rid of the chr in motifbreakr SNP_id column to conform with mpra SNP column
motif.dat$SNP <- gsub("chr","",motif.dat$SNP_id)
# Merge Motifbreakr and MPRA ( we expect a similar number to the 61570 that are in the motif.dat right now (example run not represenative of all runs))
motif.mpra.dat<-merge(motif.dat, mpra, by="SNP", all.x=T, all.y=T)
# 61579. Probably due to mpra duplicates. 
# An issue here is that there are variants in MPRA with SNP names with multiple alternate alleles, but that doesn't seem to be a huge issue. 

####### code for merging by rsid which is not used anymore #########
# Create a SNP column for the motifbreakr data (used to merge with MPRA)
# motif.dat$rsid<- motif.dat$SNP_id
# Merge Motifbreakr and MPRA
# motif.mpra.dat<-merge(motif.dat, mpra, by="rsid", all.x=T, all.y=F)

# Make sure mpra data is only the mpra_sig categories that we want
motif.mpra.dat <- subset(motif.mpra.dat, mpra_sig %in% c("Enhancer_Skew","Enhancer_nSkew","nEnhancer_nSkew"))
# Create emVar not emVar column
motif.mpra.dat$variant_skew <- ifelse(motif.mpra.dat$mpra_sig == "Enhancer_Skew", "emVar", "non_emVar")

# Make sure alleles for Motifbreakr and MPRA are the same 
# Rename the columns 
motif.mpra.dat$motifbreakr_ref <- motif.mpra.dat$REF
motif.mpra.dat$mpra_ref <- motif.mpra.dat$ref_allele
# Create a column with a 1 when the motifbreakr and mpra alleles match (Improve this loop)
motif.mpra.dat$ref_agree <- NA
motif.mpra.dat$ref_agree <- as.integer(motif.mpra.dat$motifbreakr_ref == motif.mpra.dat$mpra_ref)
# Subset to only variants in which the reference allele in the MPRA and motifbreakr match
motif.mpra.dat <- subset(motif.mpra.dat, ref_agree==1)

# Make sure alleles for Motifbreakr and MPRA are the same 
# Rename the columns 
motif.mpra.dat$motifbreakr_alt <- motif.mpra.dat$ALT
motif.mpra.dat$mpra_alt <- motif.mpra.dat$alt_allele
# Create a column with a 1 when the motifbreakr and mpra alleles match (Improve this loop)
motif.mpra.dat$alt_agree <- NA
motif.mpra.dat$alt_agree <- as.integer(motif.mpra.dat$motifbreakr_alt == motif.mpra.dat$mpra_alt)
# Subset to only variants in which the alternate allele in the MPRA and motifbreakr match
motif.mpra.dat <- subset(motif.mpra.dat, alt_agree==1)

# Add back in the variants without motifbreakr alleles
motif.mpra.dat<-merge(motif.mpra.dat, mpra, by=names(mpra), all.x=T, all.y=T)

# Generate unique MPRA sites 
motif.mpra.dat$unique_snp_tf <- paste0(motif.mpra.dat$seqnames,":",motif.mpra.dat$end,":",
                                    motif.mpra.dat$REF,":",motif.mpra.dat$ALT,"_",motif.mpra.dat$geneSymbol)
motif.mpra.dat<-motif.mpra.dat[order(-abs(motif.mpra.dat$alleleDiff)),]
motif.mpra.dat<-motif.mpra.dat[!duplicated(motif.mpra.dat$unique_snp_tf),]

write.table(motif.mpra.dat, paste0(data.dir,"motifbreakr/jurkat/", "motif.mpra.dat_",celltype_dir,"_",motif_db,".txt"), row.names=FALSE,col.names=TRUE,sep="\t", quote=F)


TF_list <-  unique(motif.mpra.dat$geneSymbol)

##################

# Plots which flip the sign of logskew for negative binders and using a binary readout for all differential TF binding

##################

# Create the TF bind or not bind data with a column for LogSkew, SNP and a column for each TF with the 

TF_all_bind_data <- subset(motif.mpra.dat, select=c(geneSymbol,LogSkew,SNP,alleleDiff))
# Pivot the data to wide format using tidyr::pivot_wider
TF_all_bind_data <- TF_all_bind_data %>%
  pivot_wider(names_from = geneSymbol, values_from = alleleDiff)

TF_all_bind_data <- unique(TF_all_bind_data)

# The filter for the number of observations needed to have sufficient sample size for each TF
TF_all_bind_data <- TF_all_bind_data[, (colSums(!is.na(TF_all_bind_data)) >= 5)==TRUE]
# If you combine this with the activity filter the results really change a lot

#For each MPRA variant, find the fold change of the allele with the largest absolute fold change
# This creates a column full of NAs for the loop to fill
mpra$max_abs_fc <- NA
# This loop calculates the column of max_abs_fc. It takes the two alleles A and B and calculates which one's absolute value is larger. That larger value is the value of the new column (can be positive or negative). 
for(i in 1:nrow(mpra)){
  if(abs(mpra[i,]$A.log2FC)>abs(mpra[i,]$B.log2FC)){
    mpra[i,]$max_abs_fc<-mpra[i,]$A.log2FC
  }else{
    mpra[i,]$max_abs_fc<-mpra[i,]$B.log2FC
  }
}

# Add back in the varaints which bind to no TFs
logskew_fc_all_vars <- subset(mpra, select=c(SNP,LogSkew,max_abs_fc))

TF_all_bind_data <- merge(logskew_fc_all_vars,TF_all_bind_data, by=c("SNP","LogSkew"), all.x=T)

write.table(TF_all_bind_data, paste0(data.dir,"motifbreakr/jurkat/", "TF_all_bind_data_",celltype_dir,"_",motif_db,".txt"), row.names=FALSE,col.names=TRUE,sep="\t", quote=F)
```

### analysis loop
Motifbreakr analysis loop for jukat cells
```{r jurkat analysis loop}
# TF list 
TF_list <- names(TF_all_bind_data)
TF_list <- TF_list[4:length(TF_list)] 

# Create test with different samples of flipping the same amount of variants in the no bind as you do in the yes bind.
# Now create a data frame with p value and effect size for all the TFs. 
t.test.all.bind.motif.dat <- NULL
hundred.p.values.per.tf <- data.frame(p.val=1:100)
# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
for(tf in TF_list){

temp_data<-TF_all_bind_data
# How many variants which bind to this particular tf are being flipped?
# Bind to TF 
specific_tf_data <- subset(temp_data,  select=c("SNP","LogSkew","max_abs_fc",tf))
# This is the sample size of the allele which are flipped for the tfs which bind to a particular TF. Change this sign as well as the one below to flip the plot
size_flipped_allelediff <- nrow(subset(specific_tf_data,  is.na(specific_tf_data[[tf]])==FALSE & specific_tf_data[[tf]]>=0))

# Find the variants which bind to a particular tf
yes_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])!=TRUE)
# flip the signs of logskew for tfs that bind and tfs that don't no bind (which we don't even nromally flip) separately. Change the sign with the one above in order to flip the plot
yes_tf_bind$LogSkew <- ifelse(!is.na(yes_tf_bind[[tf]]) & yes_tf_bind[[tf]] >= 0, -yes_tf_bind$LogSkew, yes_tf_bind$LogSkew)
# Find the tfs that do not bind to a particular tf
no_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])==TRUE)
# Find the size of rows that must be flipped by calculating the percent of rows that were flipped from the variants which do bind to a particular tf and multiplying it by the rows of the variants that do not bind 
size_flip_no_bind <- nrow(no_tf_bind)* (size_flipped_allelediff/nrow(yes_tf_bind))
# Pick a certain amount of rows to flip from those TFs which do not bind. 
rows_to_flip <- sample(1:nrow(no_tf_bind), size=size_flip_no_bind, replace=FALSE)
# Now flip the LogSkew of the rows which I picked for the variants which do not bind to that TF
no_tf_bind$LogSkew[rows_to_flip] <- -no_tf_bind$LogSkew[rows_to_flip]

# Number of variants which bind to this particular tf
n <- length(yes_tf_bind$LogSkew)
# Calculating effect size (Cohen's D)
d <- (mean(yes_tf_bind$LogSkew, na.rm = TRUE) - mean(no_tf_bind$LogSkew, na.rm = TRUE)) / (sqrt(((sd(yes_tf_bind$LogSkew, na.rm = TRUE)^2)+(sd(no_tf_bind$LogSkew, na.rm = TRUE)^2)/2)))
mu1 <- mean(yes_tf_bind$LogSkew, na.rm = TRUE)
sd1 <- sd(yes_tf_bind$LogSkew, na.rm = TRUE)
mu0 <- mean(no_tf_bind$LogSkew, na.rm = TRUE)
sd0 <- sd(no_tf_bind$LogSkew, na.rm = TRUE)
# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
skew.bind.t.test <- t.test(x=yes_tf_bind$LogSkew, y= no_tf_bind$LogSkew,alternative = c("two.sided"))
p <- skew.bind.t.test$p.value
t <- skew.bind.t.test$statistic
df <- skew.bind.t.test$parameter

t.test.all.bind.motif.dat <- rbind(t.test.all.bind.motif.dat, data.frame(tf,p,d,n,t,df,mu1,sd1,mu0,sd0))
t.test.all.bind.motif.dat

# Repeat t-test 100 times
#################################
for(i in 1:100){

# How many variants which bind to this particular tf are being flipped?
# Bind to TF 
specific_tf_data <- subset(temp_data,  select=c("SNP","LogSkew","max_abs_fc",tf))
# This is the sample size of the allele which are flipped for the tfs which bind to a particular TF. Change this sign as well as the one below to flip the plot
size_flipped_allelediff <- nrow(subset(specific_tf_data,  is.na(specific_tf_data[[tf]])==FALSE & specific_tf_data[[tf]]>=0))

# Find the variants which bind to a particular tf
yes_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])!=TRUE)
# flip the signs of logskew for tfs that bind and tfs that don't no bind (which we don't even nromally flip) separately. Change the sign with the one above in order to flip the plot
yes_tf_bind$LogSkew <- ifelse(!is.na(yes_tf_bind[[tf]]) & yes_tf_bind[[tf]] >= 0, -yes_tf_bind$LogSkew, yes_tf_bind$LogSkew)
# Find the tfs that do not bind to a particular tf
no_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])==TRUE)
# Find the size of rows that must be flipped by calculating the percent of rows that were flipped from the variants which do bind to a particular tf and multiplying it by the rows of the variants that do not bind 
size_flip_no_bind <- nrow(no_tf_bind)* (size_flipped_allelediff/nrow(yes_tf_bind))
# Pick a certain amount of rows to flip from those TFs which do not bind. 
rows_to_flip <- sample(1:nrow(no_tf_bind), size=size_flip_no_bind, replace=FALSE)
# Now flip the LogSkew of the rows which I picked for the variants which do not bind to that TF
no_tf_bind$LogSkew[rows_to_flip] <- -no_tf_bind$LogSkew[rows_to_flip]

# Calculating effect size (Cohen's D)
d <- (mean(yes_tf_bind$LogSkew, na.rm = TRUE) - mean(no_tf_bind$LogSkew, na.rm = TRUE)) / (sqrt(((sd(yes_tf_bind$LogSkew, na.rm = TRUE)^2)+(sd(no_tf_bind$LogSkew, na.rm = TRUE)^2)/2)))

# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
skew.bind.t.test <- t.test(x=yes_tf_bind$LogSkew, y= no_tf_bind$LogSkew,alternative = c("two.sided"))
p <- skew.bind.t.test$p.value
hundred.p.values.per.tf[[tf]][i] <- p
}
#################################

 }

# Get a subsetted list of tfs which only have significant p-values
sig_tfs <- data.frame(names(hundred.p.values.per.tf)[2:ncol(hundred.p.values.per.tf)])
names(sig_tfs) <- "tf"
for (i in 1:length(TF_list)) {
sig_tfs$max.p[i] <- max(hundred.p.values.per.tf[[TF_list[[i]]]])
}


t.test.all.bind.motif.dat <- merge(t.test.all.bind.motif.dat,sig_tfs,by="tf")

write.table(t.test.all.bind.motif.dat, paste0(data.dir,"motifbreakr/jurkat/", "t.test.all.bind.motif.dat_",celltype_dir,"_",motif_db,".txt"), row.names=FALSE,col.names=TRUE,sep="\t")

# Filter by genes seen in RNAseq data (hash out the line below to eliminate the RNAseq filter)
t.test.all.bind.motif.dat <- merge(t.test.all.bind.motif.dat,get(paste0(rna_cell.type,"_TPM")),by="tf")
```

### plots
Motifbreakr analysis plots for jurkat cells
```{r jurkat plots}
# Creating labels for the plot below
t.test.all.bind.motif.dat_ranked <- t.test.all.bind.motif.dat[order(t.test.all.bind.motif.dat$p),]
t.test.all.bind.motif.dat_ranked$rank <- c(1:nrow(t.test.all.bind.motif.dat_ranked))
rows <- 0.15*nrow(t.test.all.bind.motif.dat_ranked)
labels <- t.test.all.bind.motif.dat_ranked$tf[1:rows]

# Plot of t-test p-values and effect size of LogSkew comparing SNPs which bind and do not bind to TF
plot1 <- ggplot(data=t.test.all.bind.motif.dat_ranked , aes(x = d, y=(-log10(p)))) + geom_point(colour = celltype_color) + geom_text_repel(data=subset(t.test.all.bind.motif.dat_ranked , tf==labels), aes(label=tf), max.overlaps = 50) + theme_bw() + xlab("Cohen's D (effect size)") + ylab("T-test log10 p-value") + 
  labs(title="", subtitle = "LogSkew comparing SNPs which bind and do not bind to TF", caption = paste0("Cell type: ", celltype, " ; motif db: ", motif_db)) + 
  theme(panel.grid.minor = element_blank()) + geom_point(data=subset(t.test.all.bind.motif.dat_ranked, max.p>=0.05), colour = "gray50")
print(plot1)

plots.loc <- paste0(plots.dir,"motifbreakr/",celltype_dir)
plot.name <- paste0("Volcano_plot_", celltype,"_", motif_db, "_all_flipped_binding", "", ".pdf")
ggsave(plot.name,plot = plot1, path = plots.loc)


t.test.all.bind.motif.dat_ranked <- t.test.all.bind.motif.dat[order(t.test.all.bind.motif.dat$p),]
top_tf <- t.test.all.bind.motif.dat_ranked$tf[1]
top_ten_tfs <- t.test.all.bind.motif.dat_ranked$tf[1:10]


for (tf in top_ten_tfs){
  top.tf.dat <- subset(motif.mpra.dat, geneSymbol==tf, select=c(geneSymbol,alleleDiff,SNP,LogSkew,mpra_sig))
  
skew_corr <- ggplot(data=top.tf.dat, aes(x=alleleDiff,y=LogSkew)) + geom_point(color=celltype_color) + geom_smooth(method="lm",colour=celltype_color) + theme_bw() +
  labs(title = paste0("Correlation between motifbreakr allelediff and LogSkew for ", tf), 
       subtitle= paste0("Cell type: ", celltype, " ; motif db: ", motif_db), 
       caption = "TF Binding: All TF binding original LogSkew")

print(skew_corr)

plots.loc <- paste0(plots.dir,"motifbreakr/",celltype_dir)
plot.name <- paste0("Corr_allelediff_LogSkew_", tf,"_", celltype,"_", motif_db, ".pdf")
ggsave(plot.name,plot = skew_corr, path = plots.loc)
}
```

## motifbreakR enrichment analysis with each disease

### Enrichment analysis loop for all diseases

Switch back to primary t-cell labels
```{r}
cell_type <- "tcell"
motif_db <- "hocomoco"
rna_cell.type <- "tcell"
```

motifbreakR analysis loop which calculates enrichments for each disease
```{r by disease}

disease_list <- c("MS","RA","Psoriasis","T1D","IBD","UC","Crohns")

for (disease in disease_list){

# Import motif data
motif.dat <- read.table(motifbreakr_dir, sep="\t", header=T)

# Import  MPRA
mpra <- read.table(mpra.dir.1, sep="\t", header=T)
  
# subset the mpra to a particular disease
mpra <- subset(mpra, is.na(get(paste0(disease,"_pics")))==FALSE)

# Makes geneSymbol (TF name) all uppercase
motif.dat$geneSymbol <- toupper(motif.dat$geneSymbol)
# make geneSymbol (TF Name) change from slash / to underscore so that the plots can save
motif.dat$geneSymbol <- gsub("/","_",motif.dat$geneSymbol)

# Filter by genes seen in RNAseq data (hash out the line below to eliminate the RNAseq filter)
# motif.dat <- merge(motif.dat,get(paste0(rna_cell.type,"_TPM")),by="geneSymbol")


# Subset the motif data
motif.dat<-subset(motif.dat,  select=c(seqnames, end, REF, ALT,SNP_id, geneSymbol,scoreRef ,scoreAlt, alleleDiff))
# Subset MPRA
mpra <- subset(mpra, select=c(SNP,A.log2FC,B.log2FC,LogSkew,mpra_sig,rsid,ref_allele,alt_allele))

######## code for merging mpra and motifbreakr by chromosome and position ########
# Get rid of the chr in motifbreakr SNP_id column to conform with mpra SNP column
motif.dat$SNP <- gsub("chr","",motif.dat$SNP_id)
# Merge Motifbreakr and MPRA ( we expect a similar number to the 61570 that are in the motif.dat right now (example run not represenative of all runs))
motif.mpra.dat<-merge(motif.dat, mpra, by="SNP", all.x=T, all.y=F)
# 61579. Probably due to mpra duplicates. 
# An issue here is that there are variants in MPRA with SNP names with multiple alternate alleles, but that doesn't seem to be a huge issue. 

####### code for merging by rsid which is not used anymore #########
# Create a SNP column for the motifbreakr data (used to merge with MPRA)
# motif.dat$rsid<- motif.dat$SNP_id
# Merge Motifbreakr and MPRA
# motif.mpra.dat<-merge(motif.dat, mpra, by="rsid", all.x=T, all.y=F)

# Make sure mpra data is only the mpra_sig categories that we want
motif.mpra.dat <- subset(motif.mpra.dat, mpra_sig %in% c("Enhancer_Skew","Enhancer_nSkew","nEnhancer_nSkew"))
# Create emVar not emVar column
motif.mpra.dat$variant_skew <- ifelse(motif.mpra.dat$mpra_sig == "Enhancer_Skew", "emVar", "non_emVar")

# Make sure alleles for Motifbreakr and MPRA are the same 
# Rename the columns 
motif.mpra.dat$motifbreakr_alt <- motif.mpra.dat$ALT
motif.mpra.dat$mpra_alt <- motif.mpra.dat$alt_allele
# Create a column with a 1 when the motifbreakr and mpra alleles match (Improve this loop)
motif.mpra.dat$alt_agree <- NA
motif.mpra.dat$alt_agree <- as.integer(motif.mpra.dat$motifbreakr_alt == motif.mpra.dat$mpra_alt)
# Subset to only variants in which the alternate allele in the MPRA and motifbreakr match
motif.mpra.dat <- subset(motif.mpra.dat, alt_agree==1)
# length(unique(motif.mpra.dat$SNP)) # 13078 variants with rsid. 13656 with SNP. (for example run of tcells and hocomoco)
# Generate unique MPRA sites 
motif.mpra.dat$unique_snp_tf <- paste0(motif.mpra.dat$seqnames,":",motif.mpra.dat$end,":",
                                    motif.mpra.dat$REF,":",motif.mpra.dat$ALT,"_",motif.mpra.dat$geneSymbol)
motif.mpra.dat<-motif.mpra.dat[order(-abs(motif.mpra.dat$alleleDiff)),]
motif.mpra.dat<-motif.mpra.dat[!duplicated(motif.mpra.dat$unique_snp_tf),]

TF_list <-  unique(motif.mpra.dat$geneSymbol)

##################

# Plots which flip the sign of logskew for negative binders and using a binary readout for all differential TF binding

##################

# Create the TF bind or not bind data with a column for LogSkew, SNP and a column for each TF with the 

TF_all_bind_data <- subset(motif.mpra.dat, select=c(geneSymbol,LogSkew,SNP,alleleDiff))
# Pivot the data to wide format using tidyr::pivot_wider
TF_all_bind_data <- TF_all_bind_data %>%
  pivot_wider(names_from = geneSymbol, values_from = alleleDiff)

TF_all_bind_data <- unique(TF_all_bind_data)

# The filter for the number of observations needed to have sufficient sample size for each TF
TF_all_bind_data <- TF_all_bind_data[, (colSums(!is.na(TF_all_bind_data)) >= 5)==TRUE]
# If you combine this with the activity filter the results really change a lot

#For each MPRA variant, find the fold change of the allele with the largest absolute fold change
# This creates a column full of NAs for the loop to fill
mpra$max_abs_fc <- NA
# This loop calculates the column of max_abs_fc. It takes the two alleles A and B and calculates which one's absolute value is larger. That larger value is the value of the new column (can be positive or negative). 
for(i in 1:nrow(mpra)){
  if(abs(mpra[i,]$A.log2FC)>abs(mpra[i,]$B.log2FC)){
    mpra[i,]$max_abs_fc<-mpra[i,]$A.log2FC
  }else{
    mpra[i,]$max_abs_fc<-mpra[i,]$B.log2FC
  }
}

# Add back in the varaints which bind to no TFs
logskew_fc_all_vars <- subset(mpra, select=c(SNP,LogSkew,max_abs_fc))

TF_all_bind_data <- merge(logskew_fc_all_vars,TF_all_bind_data, by=c("SNP","LogSkew"), all.x=T)

# TF list 
TF_list <- names(TF_all_bind_data)
TF_list <- TF_list[4:length(TF_list)] 

# Create test with different samples of flipping the same amount of variants in the no bind as you do in the yes bind.
# Now create a data frame with p value and effect size for all the TFs. 
t.test.all.bind.motif.dat <- NULL
hundred.p.values.per.tf <- NULL
hundred.p.values.per.tf <- data.frame(p.val=1:100)
# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
for(tf in TF_list){

temp_data<-TF_all_bind_data
# How many variants which bind to this particular tf are being flipped?
# Bind to TF 
specific_tf_data <- subset(temp_data,  select=c("SNP","LogSkew","max_abs_fc",tf))
# This is the sample size of the allele which are flipped for the tfs which bind to a particular TF. Change this sign as well as the one below to flip the plot
size_flipped_allelediff <- nrow(subset(specific_tf_data,  is.na(specific_tf_data[[tf]])==FALSE & specific_tf_data[[tf]]>=0))

# Find the variants which bind to a particular tf
yes_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])!=TRUE)
# flip the signs of logskew for tfs that bind and tfs that don't no bind (which we don't even nromally flip) separately. Change the sign with the one above in order to flip the plot
yes_tf_bind$LogSkew <- ifelse(!is.na(yes_tf_bind[[tf]]) & yes_tf_bind[[tf]] >= 0, -yes_tf_bind$LogSkew, yes_tf_bind$LogSkew)
# Find the tfs that do not bind to a particular tf
no_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])==TRUE)
# Find the size of rows that must be flipped by calculating the percent of rows that were flipped from the variants which do bind to a particular tf and multiplying it by the rows of the variants that do not bind 
size_flip_no_bind <- nrow(no_tf_bind)* (size_flipped_allelediff/nrow(yes_tf_bind))
# Pick a certain amount of rows to flip from those TFs which do not bind. 
rows_to_flip <- sample(1:nrow(no_tf_bind), size=size_flip_no_bind, replace=FALSE)
# Now flip the LogSkew of the rows which I picked for the variants which do not bind to that TF
no_tf_bind$LogSkew[rows_to_flip] <- -no_tf_bind$LogSkew[rows_to_flip]

# Number of variants which bind to this particular tf
n <- length(yes_tf_bind$LogSkew)
# Calculating effect size (Cohen's D)
d <- (mean(yes_tf_bind$LogSkew, na.rm = TRUE) - mean(no_tf_bind$LogSkew, na.rm = TRUE)) / (sqrt(((sd(yes_tf_bind$LogSkew, na.rm = TRUE)^2)+(sd(no_tf_bind$LogSkew, na.rm = TRUE)^2)/2)))
mu1 <- mean(yes_tf_bind$LogSkew, na.rm = TRUE)
sd1 <- sd(yes_tf_bind$LogSkew, na.rm = TRUE)
mu0 <- mean(no_tf_bind$LogSkew, na.rm = TRUE)
sd0 <- sd(no_tf_bind$LogSkew, na.rm = TRUE)

# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
skew.bind.t.test <- t.test(x=yes_tf_bind$LogSkew, y= no_tf_bind$LogSkew,alternative = c("two.sided"))
t <- skew.bind.t.test$statistic
df <- skew.bind.t.test$parameter
p <- skew.bind.t.test$p.value

t.test.all.bind.motif.dat <- rbind(t.test.all.bind.motif.dat, data.frame(tf,p,d,n,t,df,mu1,sd1,mu0,sd0))
t.test.all.bind.motif.dat

# Repeat t-test 100 times
#################################
for(i in 1:100){

# How many variants which bind to this particular tf are being flipped?
# Bind to TF 
specific_tf_data <- subset(temp_data,  select=c("SNP","LogSkew","max_abs_fc",tf))
# This is the sample size of the allele which are flipped for the tfs which bind to a particular TF. Change this sign as well as the one below to flip the plot
size_flipped_allelediff <- nrow(subset(specific_tf_data,  is.na(specific_tf_data[[tf]])==FALSE & specific_tf_data[[tf]]>=0))

# Find the variants which bind to a particular tf
yes_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])!=TRUE)
# flip the signs of logskew for tfs that bind and tfs that don't no bind (which we don't even nromally flip) separately. Change the sign with the one above in order to flip the plot
yes_tf_bind$LogSkew <- ifelse(!is.na(yes_tf_bind[[tf]]) & yes_tf_bind[[tf]] >= 0, -yes_tf_bind$LogSkew, yes_tf_bind$LogSkew)
# Find the tfs that do not bind to a particular tf
no_tf_bind <- subset(specific_tf_data, is.na(specific_tf_data[[tf]])==TRUE)
# Find the size of rows that must be flipped by calculating the percent of rows that were flipped from the variants which do bind to a particular tf and multiplying it by the rows of the variants that do not bind 
size_flip_no_bind <- nrow(no_tf_bind)* (size_flipped_allelediff/nrow(yes_tf_bind))
# Pick a certain amount of rows to flip from those TFs which do not bind. 
rows_to_flip <- sample(1:nrow(no_tf_bind), size=size_flip_no_bind, replace=FALSE)
# Now flip the LogSkew of the rows which I picked for the variants which do not bind to that TF
no_tf_bind$LogSkew[rows_to_flip] <- -no_tf_bind$LogSkew[rows_to_flip]

# Calculating effect size (Cohen's D)
d <- (mean(yes_tf_bind$LogSkew, na.rm = TRUE) - mean(no_tf_bind$LogSkew, na.rm = TRUE)) / (sqrt(((sd(yes_tf_bind$LogSkew, na.rm = TRUE)^2)+(sd(no_tf_bind$LogSkew, na.rm = TRUE)^2)/2)))

# T-test to test whether the distribution of Allele.Diff differs between emVars and other variants 
skew.bind.t.test <- t.test(x=yes_tf_bind$LogSkew, y= no_tf_bind$LogSkew,alternative = c("two.sided"))
p <- skew.bind.t.test$p.value
hundred.p.values.per.tf[[tf]][i] <- p
}
#################################

 }


# Get a subsetted list of tfs which only have significant p-values
sig_tfs <- data.frame(names(hundred.p.values.per.tf)[2:ncol(hundred.p.values.per.tf)])
names(sig_tfs) <- "tf"
for (i in 1:length(TF_list)) {
sig_tfs$max.p[i] <- max(hundred.p.values.per.tf[[TF_list[[i]]]])
}

# Creates an adjusted p-value which is more rigorous
t.test.all.bind.motif.dat$adj.p <- p.adjust(t.test.all.bind.motif.dat$p, method = "hochberg", n = length(t.test.all.bind.motif.dat$p))

t.test.all.bind.motif.dat <- merge(t.test.all.bind.motif.dat,sig_tfs,by="tf")

# Write the table
write.table(t.test.all.bind.motif.dat, paste0(data.dir,"motifbreakr/disease_comparison/", "t.test.all.bind.motif.dat_",cell_type,"_",disease,".txt"), row.names=FALSE,col.names=TRUE,sep="\t")


# Filter by genes seen in RNAseq data (hash out the line below to eliminate the RNAseq filter)
t.test.all.bind.motif.dat <- merge(t.test.all.bind.motif.dat,get(paste0(rna_cell.type,"_TPM")),by="tf")

top_sample_size <- t.test.all.bind.motif.dat[order(-t.test.all.bind.motif.dat$n),]

if(disease=="MS"){
  key_tfs <- c("RELB", "GFI1B", "GATA3")
}
if(disease=="Crohns"){
  key_tfs <- c("ZNF85")
}
if(disease=="IBD"){
  key_tfs <- c("GFI1B", "ZNF85")
}
if(disease=="T1D"){
  key_tfs <- c("NFKB1","NFKB2","STAT3")
}
if(disease=="UC"){
  key_tfs <- c("JUN", "FOSL2")
}
if(disease=="Psoriasis"){
  key_tfs <- c("ATF1", "GFI1B", "CLOCK")
}

for(factor in key_tfs){
   sample_size <- subset(t.test.all.bind.motif.dat, tf==factor)$n
yes_bind_sample_size_hist <- ggplot(data=t.test.all.bind.motif.dat, aes(n)) + 
  geom_histogram(binwidth = 10, fill="#5D93CD") + theme_bw() +
  labs(title="Sample size of variants which bind to each tf",subtitle=paste0("Only variants in ", disease, " loci"),caption= paste0(factor, " binds to ",sample_size, " variants")) + geom_vline(xintercept = sample_size, color="#E14A4A") +
  annotate("text", x=(sample_size-2), y=50, label=factor, angle=90) + 
  annotate("text", x=(sample_size + 2), y=50, label=sample_size, angle=90)
print(yes_bind_sample_size_hist) 
}


# Lowest number of variants which bind to a particular TF
min(t.test.all.bind.motif.dat$n)

# Creating labels for the plot below
t.test.all.bind.motif.dat_ranked <- t.test.all.bind.motif.dat[order(t.test.all.bind.motif.dat$p),]
t.test.all.bind.motif.dat_ranked$rank <- c(1:nrow(t.test.all.bind.motif.dat_ranked))
rows <- 0.15*nrow(t.test.all.bind.motif.dat_ranked)
labels <- t.test.all.bind.motif.dat_ranked$tf[1:rows]

# Plot of t-test p-values and effect size of LogSkew comparing SNPs which bind and do not bind to TF
plot1 <- ggplot(data=t.test.all.bind.motif.dat_ranked , aes(x = d, y=(-log10(p)))) + geom_point(colour = "#5D93CD") + geom_text_repel(data=subset(t.test.all.bind.motif.dat_ranked , tf==labels), aes(label=tf), max.overlaps = 50) + theme_bw() + xlab("Cohen's D (effect size)") + ylab("T-test log10 p-value") + 
  labs(title=paste0(disease, " variants only"), subtitle = "LogSkew comparing SNPs which bind and do not bind to TF", caption = paste0("Cell type: ", cell_type, " ; motif db: ", motif_db)) + 
  theme(panel.grid.minor = element_blank()) + geom_point(data=subset(t.test.all.bind.motif.dat_ranked, max.p>=0.05), colour = "gray50")
print(plot1)

plots.loc <- paste0(plots.dir, "motifbreakr/disease_comparison/", cell_type,"/",disease)
plot.name <- paste0("Volcano_plot_", cell_type,"_", motif_db, "_all_flipped_binding", "_", disease, ".pdf")
 ggsave(plot.name,plot = plot1, path = plots.loc)


t.test.all.bind.motif.dat_ranked <- t.test.all.bind.motif.dat[order(t.test.all.bind.motif.dat$p),]
top_tf <- t.test.all.bind.motif.dat_ranked$tf[1]
top_ten_tfs <- t.test.all.bind.motif.dat_ranked$tf[1:10]


for (tf in top_ten_tfs){
  top.tf.dat <- subset(motif.mpra.dat, geneSymbol==tf, select=c(geneSymbol,alleleDiff,SNP,LogSkew,mpra_sig))
  
skew_corr <- ggplot(data=top.tf.dat, aes(x=alleleDiff,y=LogSkew)) + geom_point(color="#5D93CD") + geom_smooth(method="lm",colour="#5D93CD") + theme_bw() +
  labs(title = paste0("Correlation between motifbreakr allelediff and LogSkew for ", tf), 
       subtitle= paste0("Cell type: ", "cell_type", " ; motif db: ", "motif_db"), 
       caption = "TF Binding: All TF binding original LogSkew")

# print(skew_corr)

plots.loc <- paste0(plots.dir,"motifbreakr/disease_comparison/", cell_type,"/", disease)
plot.name <- paste0("Corr_allelediff_LogSkew_", tf,"_", cell_type,"_", motif_db,"_", disease, ".pdf")
ggsave(plot.name,plot = skew_corr, path = plots.loc)
}


} # End of disease loop
```

### Format data and create plots

1. Create a master data set with all the data for each disease

2. Gorgeously visualize the data with brilliant plots

```{r by disease data and plots}
####### create a master data set with all the disease p-values and effect sizes ########
 
diseases <- c("MS","RA","Psoriasis","T1D","IBD","UC","Crohns")

all.disease.t.test.data <- data.frame(tf = character(0))

for (disease in diseases) {
  # Read data 
  file_path <- paste0(data.dir,"motifbreakr/disease_comparison/", "t.test.all.bind.motif.dat_",cell_type,"_",disease,".txt")
  t.test.data <- read.table(file_path, sep="\t", header=TRUE)
  names(t.test.data) <- c("tf", paste0(disease, "_p"), paste0(disease, "_d"),paste0(disease, "_n"), paste0(disease, "_t"), paste0(disease, "_df"), paste0(disease, "_mu1"), paste0(disease, "_sd1"), paste0(disease, "_mu0"), paste0(disease, "_sd0"), paste0(disease, "_adj.p"),paste0(disease, "_max.p"))
  
  # Merge data
  if (nrow(all.disease.t.test.data) == 0) {
    all.disease.t.test.data <- t.test.data
  } else {
    all.disease.t.test.data <- merge(all.disease.t.test.data, t.test.data, by="tf", all.x=TRUE, all.y=TRUE)
  }
}

#### other type of data structure ####

diseases <- c("MS","RA","Psoriasis","T1D","IBD","UC","Crohns")

all.disease.t.test.data <- data.frame()

for (disease in diseases) {
  # Read data 
  file_path <- paste0(data.dir,"motifbreakr/disease_comparison/",
                      "t.test.all.bind.motif.dat",
                      "_",cell_type,"_", disease,".txt")
  t.test.data <- read.table(file_path, sep="\t", header=TRUE)
  names(t.test.data) <- c("tf", "p", "d","n","t",	"df",	"mu1",	"sd1",	"mu0",	"sd0", "adj_p","max.p")
  
  # Add disease column
  t.test.data$disease <- disease
  
  # Merge data
  if (nrow(all.disease.t.test.data) == 0) {
    all.disease.t.test.data <- t.test.data
  } else {
    all.disease.t.test.data <- rbind(all.disease.t.test.data, t.test.data)
  }
}

write.table(all.disease.t.test.data, paste0(data.dir,"motifbreakr/disease_comparison/", "t.test.all.bind.motif.dat_","all_disease",".txt"), row.names=FALSE,col.names=TRUE,sep="\t")

# data structure optimized by chat gpt
##################

up_top_tfs <- list()
down_top_tfs <- list()

for (z in diseases) {
  # Filter data for the current disease
  disease_data <- all.disease.t.test.data[all.disease.t.test.data$disease == z, ]

  # Sort the filtered data by p-value
  disease_data <- disease_data[order(disease_data$p), ]

  # Extract top upregulated and downregulated TFs
  up_top <- disease_data[disease_data$d >= 0, ]$tf[1:5]
  down_top <- disease_data[disease_data$d <= 0, ]$tf[1:5]

  # Store the results in lists
  up_top_tfs[[z]] <- up_top
  down_top_tfs[[z]] <- down_top
}

# Create a new data frame to store results
result_df <- data.frame(disease = rep(diseases, each = 10),
                        direction = rep(c("up", "down"), each = 5, times = 7),
                        tf = NA)

# Fill in the tf column with the extracted top TFs
for (i in 1:length(diseases)) {
  disease <- diseases[i]
  up_indices <- which(result_df$disease == disease & result_df$direction == "up")
  down_indices <- which(result_df$disease == disease & result_df$direction == "down")

  result_df$tf[up_indices] <- up_top_tfs[[disease]]
  result_df$tf[down_indices] <- down_top_tfs[[disease]]
  
}
# Assign the populated result_df to top_tfs
top_tfs <- result_df
heatmap_tf_list <- unique(top_tfs$tf)
all.disease.t.test.data.subset <- subset(all.disease.t.test.data, tf%in%heatmap_tf_list)
##############################

# Hierarchical clustering to organize the tfs in the dot plot
# From this tutorial: https://uc-r.github.io/hc_clustering
# Subset the data for only the effect size, disease and tf
hierarchical.clustering.tf.dat <- subset(all.disease.t.test.data.subset, select=c(tf,d,disease))
# Pivot so that the tfs are rows, diseases are the columns and values are the effect size
hierarchical.clustering.tf.dat <- hierarchical.clustering.tf.dat %>% pivot_wider(names_from = disease, values_from = d) %>% data.frame()
# Make the row names the tfs
rownames(hierarchical.clustering.tf.dat) <- hierarchical.clustering.tf.dat$tf
# Delete the row called tf
hierarchical.clustering.tf.dat <- subset(hierarchical.clustering.tf.dat, select=-c(tf))
# transpose the dataframe so that we can hierarchically cluster the diseases below
hierarchical.clustering.disease.dat <- data.frame(t(hierarchical.clustering.tf.dat))
# Standardize the data using scale
hierarchical.clustering.tf.dat <- scale(hierarchical.clustering.tf.dat)
# Create a dissimilarity matrix
hierarchical.clustering.tf.dat <- dist(hierarchical.clustering.tf.dat, method = "euclidean")
# Do hierarchical clustering
hierarchical.clustering.tf.dat <- hclust(hierarchical.clustering.tf.dat, method = "complete" )
# Plot the dendrogram for the clustered TFs (now a ggplot object)
# plot(hierarchical.clustering.tf.dat, cex = 0.6, hang = -1)
hierarchical.clustering.tf.dendrogram <- ggdendrogram(hierarchical.clustering.tf.dat, rotate = FALSE, theme_dendro = TRUE)
hierarchical.clustering.tf.dendrogram

ggsave(filename = paste0(cell_type,"_hierarchical.clustering.tf.dendrogram.pdf"),plot = hierarchical.clustering.tf.dendrogram, path = paste0(plots.dir,"motifbreakr/disease_comparison","/",cell_type), width=20,height=2)

# Get the list of TFs in order
tf.comparison.order <- hierarchical.clustering.tf.dat$labels[hierarchical.clustering.tf.dat$order]
tf.comparison.order

# Now hierarchically cluster the diseases
# Standardize the data using scale
hierarchical.clustering.disease.dat <- scale(hierarchical.clustering.disease.dat)
# Create a dissimilarity matrix
hierarchical.clustering.disease.dat <- dist(hierarchical.clustering.disease.dat, method = "euclidean")
# Do hierarchical clustering
hierarchical.clustering.disease.dat <- hclust(hierarchical.clustering.disease.dat, method = "complete" )
# Plot the dendrogram for the clustered diseases
# plot(hierarchical.clustering.disease.dat, cex = 0.6, hang = -1)
hierarchical.clustering.disease.dendrogram <- ggdendrogram(hierarchical.clustering.disease.dat, rotate = FALSE, theme_dendro = TRUE) + scale_y_reverse(expand = c(0.2, 0))
hierarchical.clustering.disease.dendrogram

ggsave(filename = paste0(cell_type,"_hierarchical.clustering.disease.dendrogram.pdf"),plot = hierarchical.clustering.disease.dendrogram, path = paste0(plots.dir,"motifbreakr/disease_comparison","/",cell_type), width=6,height=2)

# Get the list of diseases in order
disease.comparison.order <- hierarchical.clustering.disease.dat$labels[hierarchical.clustering.disease.dat$order]
disease.comparison.order

# Make the tfs and diseases into factors to order them in the plot
all.disease.t.test.data.subset$disease <- factor(all.disease.t.test.data.subset$disease, levels = rev(disease.comparison.order))
all.disease.t.test.data.subset$tf <- factor(all.disease.t.test.data.subset$tf, levels = tf.comparison.order)

# Create heatmap with the top 5 tfs for each disease using the effect size and p-value
ttest_d_and_p_heatmap <- ggplot(data.frame(all.disease.t.test.data.subset), aes(x = tf, y = disease)) +
  geom_point(aes(size=-log10(p), fill = d),shape = 21,color="black") + 
  xlab("tf") + ylab("disease") + labs(title="t-test p-value and effect size heatmap by disease", subtitle="tcell") + scale_fill_gradient2(low="#5D93CD",mid="white",high="#E14A4A",midpoint = 0)  +
  scale_x_discrete(guide = guide_axis(n.dodge=4)) + theme(plot.background = element_blank(),panel.background = element_blank())
ttest_d_and_p_heatmap

ggsave(filename = paste0(cell_type,"_ttest_d_and_p_heatmap.pdf"),plot = ttest_d_and_p_heatmap, path = paste0(plots.dir,"motifbreakr/disease_comparison/",cell_type), width=20,height=6)


```


## Comparison between cell types

### set up 

1. Create a data set with both primary t cell and jurkat motifbreakr enrichment values

2. Filter the data to be present in either primary t cell or jurkat RNA-seq data

```{r}
ttest.unstim.jurkat.hocomoco <- read.table(paste0(data.dir,"motifbreakr/jurkat/","t.test.all.bind.motif.dat_unstim_jurkat_hocomoco.txt"), sep="\t", header=T)
names(ttest.unstim.jurkat.hocomoco) <- c("tf", "unstim.jurkat.p", "unstim.jurkat.d", "unstim.jurkat.n", "unstim.jurkat.t","unstim.jurkat.df", "unstim.jurkat.mu1", "unstim.jurkat.sd1", "unstim.jurkat.mu0", "unstim.jurkat.sd0", "unstim.jurkat.max.p")

ttest.tcell.hocomoco <- read.table(paste0(data.dir,"motifbreakr/tcell/","t.test.all.bind.motif.dat_tcell_hocomoco.txt"), sep="\t", header=T)
names(ttest.tcell.hocomoco) <- c("tf","tcell.p","tcell.d","tcell.n", "tcell.t", "tcell.df", "tcell.mu1", "tcell.sd1", "tcell.mu0", "tcell.sd0","tcell.max.p")

# Merge the ttest datasets using 
ttest.all <- merge(ttest.unstim.jurkat.hocomoco,ttest.tcell.hocomoco, by="tf", all.y=TRUE,all.x=TRUE)

# Import RNA-seq data
# Importing dice gene expression data for primary CD4 tcells (activated)
tcell_TPM <- read.csv(paste0(data.dir,"motifbreakr/","CD4_NAIVE_TPM.csv"), header=T)
# Only the max TPM values
tcell_TPM <- subset(tcell_TPM, select=c("Additional_annotations", "X0"))
# Write the correct gene name
tcell_TPM$Additional_annotations <- gsub(";protein_coding","",tcell_TPM$Additional_annotations)
# Filter for over 1 TPM
# Rename the columns to merge with the motif data
names(tcell_TPM) <- c("tf","tcell_TPM")

jurkatTPM <- read.table(paste0(data.dir,"motifbreakr/","GSE145453_norm_counts_TPM_GRCh38.p13_NCBI.tsv"), sep = '\t', header=T)
annotateTPM <- read_excel(paste0(data.dir,"motifbreakr/","idmap.xlsx"))
annotateTPM <- subset(annotateTPM, select=c(query,symbol))
names(annotateTPM) <- c("GeneID","tf")
jurkatTPM <- merge(jurkatTPM,annotateTPM, by="GeneID")
jurkatTPM$unstim_jurkat_TPM <- pmax(jurkatTPM$GSM4318505,jurkatTPM$GSM4318506)
jurkatTPM$stim_jurkat_TPM <- pmax(jurkatTPM$GSM4318507,jurkatTPM$GSM4318508)
unstim_jurkat_TPM<- subset(jurkatTPM, select=c(tf,unstim_jurkat_TPM))
stim_jurkat_TPM<- subset(jurkatTPM, select=c(tf,stim_jurkat_TPM))

# APPLY RNA-seq data
# Merge the rna-seq data for each cell type to the ttest data for all the cell types
ttest.all.tf.TPM <- merge(ttest.all, tcell_TPM, by="tf", all.x=TRUE)
ttest.all.tf.TPM <- merge(ttest.all.tf.TPM, unstim_jurkat_TPM, by="tf", all.x=TRUE)
#ttest.all.tf.TPM <- merge(ttest.all.tf.TPM, stim_jurkat_TPM, by="tf", all.x=TRUE)
# Subset the data so tfs in the data need to be listed in the data for all cell types (making this "or" in stead of "and" only adds 1 TF (BMAL1) which is NA in the tcell data)
ttest.all.tf.TPM.filtered <- subset(ttest.all.tf.TPM, is.na(tcell_TPM)==FALSE & is.na(unstim_jurkat_TPM)==FALSE & is.na(stim_jurkat_TPM)==FALSE)
# Subset the data so tfs in the data expressed in at least one cell type (making this "or" instead of "and" makes the number of tfs decrease the number of tfs)
ttest.all.tf.TPM.filtered.twice <- subset(ttest.all.tf.TPM.filtered, tcell_TPM>=1 | unstim_jurkat_TPM>=1)
```

### plots

Create plots to compare the p-values and effect sizes for tfs between primary tcells and unstimulated jurkat
```{r}
max(-log10(ttest.all.tf.TPM.filtered.twice$unstim.jurkat.p))
max(-log10(ttest.all.tf.TPM.filtered.twice$tcell.p))

# #5D93CD #E14A4A #5D93FF
multicolor.ttest.d.tcell.unstim.jurkat.compare2 <- ggplot(data=ttest.all.tf.TPM.filtered.twice,aes(x=unstim.jurkat.d,y=tcell.d)) +
 geom_hline(yintercept=0, color="black") + geom_vline(xintercept = 0, color="black") +
  geom_point(aes(color=-log10(tcell.p),fill=-log10(unstim.jurkat.p)),size=3, shape = 21,stroke=1.5) +
  scale_color_gradientn(colors=c("gray90","#E14A4A"),limits = c(0, 5.75)) +
  # scale_size_continuous(range = c(2, 6)) +
  scale_fill_gradientn(colors=c("gray90","#5D93CD"),limits = c(0, 5.75)) +
  theme_bw() + geom_text_repel(aes(label=tf)) + 
  labs(title ="Unstimulated Jurkat and Primary Tcell comparison", subtitle= "p-values and effect size derived from t-test of LogSkew and motifbreakr binding", caption="Hocomoco motif db", x ="Unstimulated Jurkat Effect size (d)", y = "Primary Tcell effect size (d)")
multicolor.ttest.d.tcell.unstim.jurkat.compare2

ggsave("multicolor2.rings2.effect.size.compare.plot.tcell.unstim.jurkat.pdf",plot = multicolor.ttest.d.tcell.unstim.jurkat.compare2, path = paste0(plots.dir,"motifbreakr"), width=7.5,height=5)
```

How many emVars disrupt key TFs?
```{r}
tcell_motifbreakr <- read.table(paste0(data.dir,"motifbreakr/tcell/","motif.mpra.dat_tcell_hocomoco.txt"), header=T, sep="\t")

key_tfs <- c("NFKB1", "NFKB2", "RELB", "FOSB", "JUN", "GATA3", "STAT3", "STAT4","GFI1B")
for(factor in key_tfs) {
  single_tf <- subset(tcell_motifbreakr, geneSymbol==factor)
n <- nrow(single_tf)
emvars <- nrow(subset(single_tf, variant_skew=="emVar"))
print(paste0("Out of a total of ",n, " variants which bind to ",factor, ", ", emvars, " are emVars."))
}
```


